---
title: "R- Markdown /NYC Taxi trip Analysis"
output: html_document
---

#LOADING DATA
```{r}
Mar17_data = read.csv('yellow_tripdata_2017-03.csv')
```

```{r}
Jun17_data = read.csv('yellow_tripdata_2017-06.csv')
```

```{r}
Nov17_data = read.csv('yellow_tripdata_2017-11.csv')
```

#MARCH DATA EXPLORATION
```{r}
head(Mar17_data)
```

```{r}
tail(Mar17_data)
```

```{r}
mean(Mar17_data$tip_amount)
```

```{r}
median(Mar17_data$tip_amount)
```

```{r}
mean(Mar17_data$trip_distance)
```

```{r}
median(Mar17_data$trip_distance)
```

```{r}
calculate_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
calculate_mode(Mar17_data$tip_amount)
```
```{r}
calculate_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
calculate_mode(Mar17_data$trip_distance)
```
```{r}
colnames(Mar17_data)
```

#Any missing values?

```{r}
sum(is.na(Mar17_data))
```
```{r}
sum(is.na(Jun17_data))
```
```{r}
sum(is.na(Nov17_data))
```


```{r}
range(Mar17_data$tip_amount)
```
```{r}
boxplot(Mar17_data$total_amount)
```

```{r}
boxplot(Mar17_data$tip_amount)
```


Negative tip amount does not make any sense. we can just either take them out of our analysis or impute them. I choose to take them out as they seem to be incorrectly inputed data.  <---- OUTLIERS

```{r}
Mar17_data= Mar17_data[!Mar17_data$total_amount < 0,]
```


```{r}
range(Mar17_data$tip_amount)
```

```{r}
Mar17_data= Mar17_data[!Mar17_data$tip_amount < 0,]
```

```{r}
range(Mar17_data$tip_amount)
```

```{r}
range(Mar17_data$total_amount)
```

Huge range for a taxi fare, Let's examine all outliers not just negative ones.

```{r}
outliers <- boxplot(Mar17_data$total_amount, plot=FALSE)$out
```

```{r}
boxplot(Mar17_data$total_amount, plot=FALSE)$out
```

We can see there is 944,805 observations that are considered outliers. Again we can either impute them with a value such as average or remove them. 
```{r}
NROW(outliers)/nrow(Mar17_data)
```

About 9% of the data are outliers, it is risky to exclude them from our analysis and many of them are resonable fare for longer distance or other circumstances, etc.


```{r}
boxplot(Mar17_data$total_amount)
```

```{r}
x <- Mar17_data$total_amount
qnt <- quantile(x, probs=c(.25, .75), na.rm = T)
caps <- quantile(x, probs=c(.05, .95), na.rm = T)
H <- 1.5 * IQR(x, na.rm = T)
x[x < (qnt[1] - H)] <- caps[1]
x[x > (qnt[2] + H)] <- caps[2]
```


```{r}
boxplot(x)
```





```{r}
Mar17_data$total_amount <- x
```


```{r}
boxplot(Mar17_data$tip_amount)
```

```{r}
z <- Mar17_data$tip_amount
qnt <- quantile(z, probs=c(.25, .75), na.rm = T)
caps <- quantile(z, probs=c(.05, .95), na.rm = T)
H1 <- 1.5 * IQR(x, na.rm = T)
z[z < (qnt[1] - H1)] <- caps[1]
z[z > (qnt[2] + H1)] <- caps[2]
```

```{r}
Mar17_data$tip_amount <- z
```

```{r}
boxplot(z)
```


```{r}

library(caTools)
set.seed(100)
split = sample.split(Mar17_data$tip_amount,SplitRatio = 0.7)
train = Mar17_data[split,]
test = Mar17_data[!split,]
model1 = glm(tip_amount~.,data=train,family='binomial')
summary(model1)
```

